---
title: "project_benoit_clemenceau"
output: html_document
---

```{r}
if (!require(dplyr)) {
  install.packages("dplyr")
  library(dplyr)
}
if (!require(ggplot2)) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require(tidyverse)) {
  install.packages("tidyverse")
  library(tidyverse)
}
if (!require(broom)) {
  install.packages("broom")
  library(broom)
}
if (!require(GGally)) {
  install.packages("GGally")
  library(GGally)
}
if (!require(aod)) {
  install.packages("aod")
  library(aod)
}
if (!require(pROC)) {
  install.packages("pROC")
  library(pROC)
}
```

```{r}
titanic = read.csv(file = "titanic/train.csv")
head(titanic)
summary(titanic)
nrow(titanic)
```

We have 12 columns in this data set that we will describe one by one.  
We also have 891 rows which is more than the 500 observations recommended in the project instructions. We will see how many rows we have left after cleaning up the data.

# Data set description

## PassengerId

```{r}
print("PassengerId")
head(titanic$PassengerId)
paste("Type:", class(titanic$PassengerId))
paste("Distincts:", n_distinct(titanic$PassengerId))
paste("NAs:", sum(is.na(titanic$PassengerId) == TRUE))
# table(titanic$PassengerId)
summary(titanic$PassengerId)
```

PassengerId is only an index for the rows.  
It will probably not be helpful for the regression, but we should keep it for context.

## Survived

```{r}
print("Survived")
head(titanic$Survived)
paste("Type:", class(titanic$Survived))
paste("Distincts:", n_distinct(titanic$Survived))
paste("NAs:", sum(is.na(titanic$Survived) == TRUE))
table(titanic$Survived)
summary(titanic$Survived)
```

Survived will be our response variable.  
It is a categorical binary (0 or 1) and has no NAs.  
From the data set information: 0 = No, 1 = Yes.  
We have a mean of 0.3838 which means that on average 38% of people survived.

## Pclass

```{r}
print("Pclass")
head(titanic$Pclass)
paste("Type:", class(titanic$Pclass))
paste("Distincts:", n_distinct(titanic$Pclass))
paste("NAs:", sum(is.na(titanic$Pclass) == TRUE))
table(titanic$Pclass)
summary(titanic$Pclass)
```

Pclass is the ticket class.  
It is a categorical variable with 3 options: 1, 2, 3.  
We can see that most people were in 3rd class, then 1st and 2nd class were roughly as popular, with 1st class being slighlty more.  
This should be a useful variable for the regression.

## Name

```{r}
print("Name")
head(titanic$Name)
paste("Type:", class(titanic$Name))
paste("Distincts:", n_distinct(titanic$Name))
paste("NAs:", sum(is.na(titanic$Name) == TRUE))
# table(titanic$Name)
summary(titanic$Name)
```

Name of the passengers.  
891 distinct values, no NAs.  
Won't be useful for the regression except if we do some kind of feature extraction on it.  

## Sex

```{r}
print("Sex")
head(titanic$Sex)
paste("Type:", class(titanic$Sex))
paste("Distincts:", n_distinct(titanic$Sex))
paste("NAs:", sum(is.na(titanic$Sex) == TRUE))
table(titanic$Sex)
summary(titanic$Sex)
```

Sex of the passengers.  
2 distinct values, no NAs.  
More males than females.  
Should be converted to a binary categorical variable.  
Will probably be very useful to the regression.  

## Age

```{r}
print("Age")
head(titanic$Age)
paste("Type:", class(titanic$Age))
paste("Distincts:", n_distinct(titanic$Age))
paste("NAs:", sum(is.na(titanic$Age) == TRUE))
# table(titanic$Age)
summary(titanic$Age)
```

Age of the passengers.  
89 distinct values, lots of NAs.  
Will need to do something about the missing values.  
Will probably be very useful to the regression.  

## SibSp

```{r}
print("SibSp")
head(titanic$SibSp)
paste("Type:", class(titanic$SibSp))
paste("Distincts:", n_distinct(titanic$SibSp))
paste("NAs:", sum(is.na(titanic$SibSp) == TRUE))
table(titanic$SibSp)
summary(titanic$SibSp)
```

Number of siblings / spouses aboard the Titanic.  
7 distinct values, no NAs.  
Will probably be somewhat useful to the regression.  

## Parch

```{r}
print("Parch")
head(titanic$Parch)
paste("Type:", class(titanic$Parch))
paste("Distincts:", n_distinct(titanic$Parch))
paste("NAs:", sum(is.na(titanic$Parch) == TRUE))
table(titanic$Parch)
summary(titanic$Parch)
```

Number of parents / children aboard the Titanic.  
7 distinct values, no NAs.  
Will probably be very useful to the regression.  

## Ticket

```{r}
print("Ticket")
head(titanic$Ticket)
paste("Type:", class(titanic$Ticket))
paste("Distincts:", n_distinct(titanic$Ticket))
paste("NAs:", sum(is.na(titanic$Ticket) == TRUE))
# table(titanic$Ticket)
summary(titanic$Ticket)
```

Ticket number.  
681 distinct values, no NAs.  
Won't be useful for the regression unless we do some kind of feature extraction on it.  

## Fare

```{r}
print("Fare")
head(titanic$Fare)
paste("Type:", class(titanic$Fare))
paste("Distincts:", n_distinct(titanic$Fare))
paste("NAs:", sum(is.na(titanic$Fare) == TRUE))
# table(titanic$Fare)
summary(titanic$Fare)
```

Passenger fare.  
248 distinct values, no NAs.  
Could be useful for the regression but certainly highly correlated to Pclass.  

## Cabin

```{r}
print("Cabin")
head(titanic$Cabin)
paste("Type:", class(titanic$Cabin))
paste("Distincts:", n_distinct(titanic$Cabin))
paste("NAs:", sum(is.na(titanic$Cabin) == TRUE))
# table(titanic$Cabin)
summary(titanic$Cabin)
```

Cabin number.  
148 distinct values, lots of NAs.  
Probably won't be useful for the regression unless we do some kind of feature extraction on it.  

## Embarked

```{r}
print("Embarked")
head(titanic$Embarked)
paste("Type:", class(titanic$Embarked))
paste("Distincts:", n_distinct(titanic$Embarked))
paste("NAs:", sum(is.na(titanic$Embarked) == TRUE))
table(titanic$Embarked)
summary(titanic$Embarked)
```

Port of Embarkation.  
C = Cherbourg, Q = Queenstown, S = Southampton.
4 distinct values, 0 NAs, 2 blank values.  
Could be useful for the regression.  

# Cleaning up the data

## Survived

Converting to factor.

```{r}
titanic$Survived = factor(titanic$Survived)
```

## Pclass

Converting to factor.

```{r}
titanic$Pclass = factor(titanic$Pclass)
```

## Name

We drop this column as of now because each row has a different string value.

```{r}
titanic = subset(titanic, select = -c(Name))
```

## Sex

Converting to factor.

```{r}
titanic$Sex = factor(titanic$Sex, labels = c("F", "M"))
```

## Age

Dropping rows with missing values as we already have too much rows.

```{r}
titanic = titanic[complete.cases(titanic[, c("Age")]), ]
```

Looking at the distribution:

```{r fig.align="center", fig.width = 8, fig.height = 4}
ggplot(titanic,
       aes(x = Age)) +
    geom_histogram(binwidth = 2.5, fill = "blue", color = "blue", alpha = .5) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
    theme_bw() +
    ggtitle("Age distribution") +
    xlab("Age") +
    ylab("Count")
```

It is skewed to the right and there might be outliers.

```{r fig.align="center", fig.width = 2, fig.height = 4}
ggplot(titanic) +
  aes(x = "", y = Age) +
  geom_boxplot() +
  theme_bw()
```

Removing outliers:

```{r}
quartiles = quantile(titanic$Age, probs = c(.25, .75), na.rm = FALSE)
upper = quartiles[2] + 1.5 * IQR(titanic$Age)
titanic = subset(titanic, titanic$Age <= upper)
```

```{r fig.align="center", fig.width = 8, fig.height = 4}
ggplot(titanic,
       aes(x = Age)) +
    geom_histogram(binwidth = 2.5, fill = "blue", color = "blue", alpha = .5) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
    theme_bw() +
    ggtitle("Age distribution") +
    xlab("Age") +
    ylab("Count")
```

```{r fig.align="center", fig.width = 2, fig.height = 4}
ggplot(titanic) +
  aes(x = "", y = Age) +
  geom_boxplot() +
  theme_bw()
```

## SibSp

Converting to factors:

```{r}
titanic$SibSp = factor(titanic$SibSp)
```

## Parch

Converting to factors:

```{r}
titanic$Parch = factor(titanic$Parch)
```

## Ticket

We drop this column as of now.

```{r}
titanic = subset(titanic, select = -c(Ticket))
```

## Fare

Might be highly correlated to Pclass, we will run tests after this part to decide what we do with it.

Looking at the distribution:

```{r fig.align="center", fig.width = 8, fig.height = 4}
ggplot(titanic,
       aes(x = Fare)) +
    geom_histogram(binwidth = 2.5, fill = "blue", color = "blue", alpha = .5) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
    theme_bw() +
    ggtitle("Fare distribution") +
    xlab("Fare") +
    ylab("Count")
```

It is heavily skewed to the right and there are a lot of outliers.

```{r fig.align="center", fig.width = 2, fig.height = 4}
ggplot(titanic) +
  aes(x = "", y = Fare) +
  geom_boxplot() +
  theme_bw()
```

Log transform:

```{r}
titanic$Fare = log(titanic$Fare + 1)
```

```{r fig.align="center", fig.width = 8, fig.height = 4}
ggplot(titanic,
       aes(x = Fare)) +
    geom_histogram(binwidth = 0.25, fill = "blue", color = "blue", alpha = .5) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
    theme_bw() +
    ggtitle("Fare distribution") +
    xlab("log(Fare)") +
    ylab("Count")
```

```{r fig.align="center", fig.width = 2, fig.height = 4}
ggplot(titanic) +
  aes(x = "", y = Fare) +
  geom_boxplot() +
  theme_bw() +
  ylab("log(Fare)")
```

Removing outliers:

```{r}
quartiles = quantile(titanic$Fare, probs = c(.25, .75), na.rm = FALSE)
iqr = IQR(titanic$Fare)
lower = quartiles[1] - 1.5 * iqr
upper = quartiles[2] + 1.5 * iqr
titanic = subset(titanic, titanic$Fare >= lower & titanic$Fare <= upper)
```

```{r fig.align="center", fig.width = 8, fig.height = 4}
ggplot(titanic,
       aes(x = Fare)) +
    geom_histogram(binwidth = 0.25, fill = "blue", color = "blue", alpha = .5) +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
    theme_bw() +
    ggtitle("Fare distribution") +
    xlab("log(Fare)") +
    ylab("Count")
```

```{r fig.align="center", fig.width = 2, fig.height = 4}
ggplot(titanic) +
  aes(x = "", y = Fare) +
  geom_boxplot() +
  theme_bw() +
  ylab("log(Fare)")
```


## Cabin

We drop this column as of now.

```{r}
titanic = subset(titanic, select = -c(Cabin))
```

## Embarked

Dropping rows with blank values and converting to factor.

```{r}
titanic$Embarked[titanic$Embarked == ""] = NA
titanic = titanic[complete.cases(titanic[, c('Embarked')]), ]
titanic$Embarked = factor(titanic$Embarked)
```

# Cleaned data set

```{r}
head(titanic)
summary(titanic)
n = nrow(titanic)
nrow(titanic)
```

We still have too many rows to comply with the instructions' limit of 500. Let's delete some randomly:

```{r}
titanic = titanic[-sample(1:nrow(titanic), n - 500), ]
summary(titanic)
nrow(titanic)
```

# Building the model

```{r}
mdl = glm(
  Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked,
  family = binomial,
  data = titanic
)
```

# Checking logistic regression assumptions

## Dependent variable is binary

```{r}
table(titanic$Survived)
```
 
TRUE
 
## Observations are independent of one another

This is hard to say as we don't have a way of testing it.  
We assume that it is true.

## Little or no multicollinearity among the independent variables

```{r}
ggpairs(subset(titanic, select = -c(PassengerId)))
```

Looks OK.

## Independent variables are linearly associated with the log odds of the outcome

```{r}
# We run it on continuous predictors only
continuous_predictors = subset(titanic, select = c(Age, Fare))
predictors_names = colnames(continuous_predictors)

probabilities = predict(mdl, type = "response")

continuous_predictors = continuous_predictors %>%
  mutate(logit = log(probabilities / (1 - probabilities))) %>%
  gather(key = "predictors", value = "predictor.value",-logit)
```

```{r}
ggplot(continuous_predictors, aes(logit, predictor.value))+
  geom_point(size = 0.5, alpha = 0.5) +
  geom_smooth(method = "loess") + 
  theme_bw() + 
  facet_wrap(~predictors, scales = "free_y")
```

Not at all, maybe we should remove these variables from the model.

```{r}
mdl = glm(
  Survived ~ Pclass + Sex + SibSp + Parch + Embarked,
  family = binomial,
  data = titanic
)
```

## Enough samples

The rule of thumb is 10 cases with the least frequent outcome for each independent variable in the model.
We need 10 x 5 = 50 for the least frequent outcome.

```{r}
table(titanic$Survived)
```

TRUE

# Evaluating the model

```{r}
wald.test(b = coef(mdl), Sigma = vcov(mdl), Terms = 2:6)
```

Reject $H_0$ since $\text{overall p-value} = 0.0 ≤ \alpha = 0.05$.  
We have significant evidence at the $\alpha = 0.05$ level that at least one $\beta_i ≠ 0$.

```{r}
summary(mdl)
```

$\beta_{Pclass1} = 2.8127$ ($\text{p-value} = 1.16e-08 ≤ \alpha = 0.05$)  
$\beta_{Pclass2} = -1.2984$ ($\text{p-value} = 0.000702 ≤ \alpha = 0.05$)  
$\beta_{Pclass3} = -1.9963$ ($\text{p-value} = 1.59e-07 ≤ \alpha = 0.05$)  
$\beta_{SexM} = -3.1176$ ($\text{p-value} = < 2e-16 ≤ \alpha = 0.05$)

We are unable to say that all the other coefficients are significant.

```{r fig.align="center", fig.width = 5, fig.height = 5}
titanic$prob = predict(mdl, type = c("response"))
plot(roc(Survived ~ prob, data = titanic),
     print.auc = TRUE,
     auc.polygon = TRUE)
```

The c-statistic of the model is 0.848 (value may differ slightly because of the random sampling of the data to have 500 observations).

```{r}
titanic$pred = ifelse(titanic$prob >= 0.50, 1, 0)

# Confusion matrix
table(titanic$Survived, titanic$pred)

error = mean(titanic$Survived != titanic$pred)
paste("Accuracy:", round(1 - error, 4))
```

**We get an accuracy of 79% with this model when choosing a cutoff of 0.5**  
(value may differ slightly because of the random sampling of the data to have 500 observations)

# Second model with only Pclass and Sex as predictors

```{r}
mdl_2 = glm(
  Survived ~ Pclass + Sex,
  family = binomial,
  data = titanic
)
```

```{r}
wald.test(b = coef(mdl_2), Sigma = vcov(mdl_2), Terms = 2:3)
```

Reject $H_0$ since $\text{overall p-value} = 6.3e-10 ≤ \alpha = 0.05$.  
We have significant evidence at the $\alpha = 0.05$ level that at least one $\beta_i ≠ 0$.

```{r}
summary(mdl_2)
```

$\beta_{Pclass1} = 2.3174$ ($\text{p-value} = 7.94e-16 ≤ \alpha = 0.05$)  
$\beta_{Pclass2} = -1.0760$ ($\text{p-value} = 0.000599 ≤ \alpha = 0.05$)  
$\beta_{Pclass3} = -1.8393$ ($\text{p-value} = 8.30e-11 ≤ \alpha = 0.05$)  
$\beta_{SexM} = -2.5041$ ($\text{p-value} = < 2e-16 ≤ \alpha = 0.05$)

```{r fig.align="center", fig.width = 5, fig.height = 5}
titanic$prob_2 = predict(mdl_2, type = c("response"))
plot(roc(Survived ~ prob_2, data = titanic),
     print.auc = TRUE,
     auc.polygon = TRUE)
```

The c-statistic of the model is 0.821. (value may differ slightly because of the random sampling of the data to have 500 observations)

```{r}
titanic$pred_2 = ifelse(titanic$prob_2 >= 0.50, 1, 0)

# Confusion matrix
table(titanic$Survived, titanic$pred_2)

error = mean(titanic$Survived != titanic$pred_2)
paste("Accuracy:", round(1 - error, 4))
```

**We get an accuracy of 78.6% using only Pclass and Sex as predictors when choosing a cutoff of 0.5**  
(value may differ slightly because of the random sampling of the data to have 500 observations)

# Exporting the final dataset

```{r}
write.csv(subset(titanic, select = c(Pclass, Sex, SibSp, Parch, Embarked, Survived)), "project_benoit_clemenceau.csv", row.names = TRUE)
```


